Predicting Severity of _C. difficile_ Infections from the Taxonomic Composition of the Gut Microbiome

### 2023-06-07 paper/paper.qmd
Copy-paste supplement instead of using 'include'

title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"

@@

@@

  We investigated whether machine learning (ML) models trained on gut microbiota
  isolated from stool samples on the day of CDI diagnosis could predict which
  CDI cases led to severe outcomes.
  We collected 1,277 stool samples from CDI patients and clustered 16S rRNA gene
  amplicon sequences into _de novo_ Operational Taxonomic Units (OTUs).

@@

  and falls back to the all-cause definition otherwise.

@@

  model performance. 

@@

_Clostridoides difficile_ infection (CDI) is the most common nosocomial
infection in the United States, and community-acquired cases are on the rise
[@magill_multistate_2014;@feuerstadt_burden_2023].
The classic CDI case typically occurs soon after antibiotic use, which perturbs the
protective gut microbiota and allows _C. difficile_ to proliferate [@kelly_can_2012].
Non-antibiotic medications including proton-pump inhibitors and osmotic laxatives
have also been shown to increase CDI susceptibilty and inhibit clearance
[@janarthanan_clostridium_2012;@tomkovich_osmotic_2021].
Diarrhea is the primary symptom, with some patients developing colitis, 
toxic megacolon, or requiring intensive care with an in-hospital mortality rate
of approximately 8-9% [@lucado_clostridium_2012;@kassam_clostridium_2016].
Furthermore, 5-20% of initial cases reoccur within 2-8 weeks, and recurrent
cases are associated with increased morbidity and mortality risk
[@napolitano_clostridium_2017;@kelly_can_2012].
Patient risk factors for CDI-related morbidity and mortality include age > 65,
history of recurrent CDI, and co-morbid chronic illnesses [@ressler_defining_2021].
CDI remains a significant burden on the US health care system with approximately
500,000 cases annually [@guh_trends_2020;@kwon_morbidity_2015].

There is a need for robust, accurate methods to identify patients at risk of
severe CDI outcomes.
When paired with treatment options that may reduce risk of severity, prediction
models can guide clinician decision-making to improve patient outcomes while
minimizing harms from unnecessary treatment.
Numerous scoring systems for predicting severe CDI outcomes based on patient
clinical factors have been developed, but none have validated well to external
datasets nor are any in use in routine clinical practice
[@chakra_prediction_2012;@perry_external_2022].
Machine learning (ML) is a promising approach that allows for use of thousands
of features to classify samples and predict outcomes, rather than relying on
limited sets of human-curated variables.
Indeed, ML models trained on entire Electronic Health Record (EHR) data
have demonstrated improved performance over curated models
[@rao_clostridium_2015;@li_using_2019].

@@

## CDI severity

There is not currently a consensus definition of CDI severity.
Some scoring systems leverage clinical data available during the course of CDI,
while others focus on adverse outcomes of CDI at 30 days after diagnosis
[@ressler_defining_2021;@dieterle_systemic_2020].
We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
The "IDSA" definition of severe CDI is based on laboratory values collected on
the day of diagnosis, with a case being severe if serum creatinine level is
greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
However, the IDSA score is known to be a poor predictor of adverse outcomes
[@stevens_validation_2020], although it is straightforward to collect.
The remaining definitions we employed focus on the occurrence of adverse
outcomes, which may be more clinically relevant.
The "attributable" severity definition is based on disease-related complications
defined by the CDC, where an adverse event of ICU admission, colectomy, or death
occurs within 30 days of CDI diagnosis, and the adverse event is determined to
be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
However, physician chart review is time-consuming and has not been completed for
all cases, so we defined "all-cause" severity where a case is severe if an
adverse event occurs within 30 days of the diagnosis regardless of the cause of
the adverse event.
Finally, we defined a "pragmatic" severity definition that makes use of the
attributable definition when available and falls back to the all-cause
definition when chart review has not been completed, allowing us to use as many
samples as we have available while taking physicians' expert opinions into
account where possible (@fig-flowchart B).
We trained ML models to classify (in the case of the IDSA definition) or predict
(in the case of the three other definitions) severity and determined how well
OTU-based models perform for each definition.

@@

```{r sample_counts}
#| message: false
#| label: tbl-counts
#| tbl-pos: h
#| tbl-cap: "**Sample counts and proportion of severe cases.** Each severity 
#|   definition has a different number of patient samples available, as well as
#|   a different proportion of cases labelled as severe.
#| "
#| tbl-subcap: 
#|   - "Full datasets"
#|   - "Intersection of samples with all labels available"
#| layout-ncol: 2
format_table <- function(tab) {
  tab %>%
    kable(col.names = c(' ', 'IDSA', 'All-cause', 'Attributable', "Pragmatic"),
          format.args = list(big.mark = ',')) %>% 
    kable_styling(font_size = table_font_size)
}
read_csv(here('results','count_table_full.csv')) %>% 
  format_table()
read_csv(here('results','count_table_int.csv')) %>% 
  format_table()
```

## Model performance

We first set out to train the best models possible for each severity definition.
Not all samples have labels available for all four severity definitions due to
missing data for some patient lab values and incomplete chart review
(@fig-flowchart B), thus each severity definition has a different number of
samples when using as many samples as possible (@tbl-counts A).
We refer to these as the full datasets.
Random forest models were trained on 100 splits of the datasets into training
and test sets, and performance was evaluated on the held-out test set using the
area under the receiver-operator characteristic curve (AUROC).
Since the severity classes are highly imbalanced with different proportions of
severe samples between definitions, we also calculated the balanced precision
and the area under the balanced precision-recall curve (AUBPRC) as first
proposed by @wu_improved_2021 to describe the precision that would be expected
if the outcome classes were balanced.

@@

After training on the full datasets, the performance as measured by the AUROCs
of the training set cross-validation folds were similar to those of the held-out
test sets, indicating that the models are neither overfit nor underfit
(@fig-performance A).
As measured by AUROC on the held-out test sets, models predicting pragmatic
severity performed best with a median AUROC of `r perf_medians$auroc_full_pragmatic`,
and this was significantly different from that of the other definitions 
on the full datasets (P < 0.05).
Models predicting IDSA, all-cause, and attributable severity performed similarly
with median test set AUROCs of
`r perf_medians$auroc_full_idsa`,
`r perf_medians$auroc_full_allcause`, and
`r perf_medians$auroc_full_attrib` respectively.
The test set AUROCs were not significantly different (P > 0.05) for attributable
and IDSA nor for attributable and all-cause, but the IDSA and all-cause AUROCs
were significantly different from each other (P < 0.05).
We plotted the receiver-operator characteristic curve and found that the
pragmatic severity models outperformed the others at all specificity values
(@fig-performance B).
For comparison, a prior study with a different dataset trained a logistic
regression model on whole Electronic Health Record data extracted on the day of
CDI diagnosis to predict attributable severity, yielding an AUROC of `r
ehr_auroc` [@li_using_2019].
While our OTU-based attributable severity model did not meet this performance,
the OTU-based pragmatic severity model performed just as well as the EHR-based
model in terms of AUROC.

::: {#fig-performance fig-cap="**Performance of ML models.**
In the left facets, models were trained on the full datasets, with different
numbers of samples available for each severity definition.
In the right facets, models were trained on the same dataset consisting of the
intersection of samples with labels available for all definitions.
Note that the intersection dataset has exactly the same labels for attributable
and pragmatic severity, thus these have identical performance.
**A)** 
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets, and the area under the balanced
precision-recall curve (AUBPRC) for the test sets.
Each point is annotated with the median performance across 100 train/test splits
with tails as the 95% CI.
**B)** Receiver-operator characteristic curves for the test sets.
Mean specificity is reported at each sensitivity value, 
with ribbons as the 95% CI.
**C)** Balanced precision-recall curves for the test sets.
Mean balanced precision is reported at each recall (sensitivity) value, 
with ribbons as the 95% CI.
Original unbalanced precision-recall curves are shown in Figure S1.
"}
![](figures/ml-performance.png)
:::

@@

## Feature importance

@@

## Estimating clinical value

Even if a model performs well, it may not be useful in a clinical setting unless
it can guide clinicians to choose between treatment options.
At this time, we are not aware of any direct evidence that a particular
treatment reduces the risk of severe CDI outcomes.
However, with some assumptions we offer a proof-of-concept analysis of the
potential clinical value of OTU-based severity prediction models when paired
with treatments that may reduce severity.
When considering the suitability of a model for deployment in clinical settings,
the number needed to screen (NNS) is a highly relevant metric representing how
many patients must be predicted as severe by the model to identify one true
positive.
Similarly, the number needed to treat (NNT) is the number of true positive
patients that must be treated by an intervention in order for one patient to
benefit from the treatment.
Multiplying NNS by NNT yields the number needed to benefit (NNB): the number
of patients predicted to have a severe outcome who then benefit from the
treatment [@liu_number_2019].
Thus the NNB pairs model performance with treatment effectiveness to estimate
the benefit of using predictive models in clinical practice, and are useful for
comparing models and performing cost-benefit analyses.

@@

To calculate a clinically-relevant NNS for these models, we computed the
confusion matrix at the 95th percentile of risk for each prediction model.
We excluded the IDSA severity models as the IDSA severity scores were calculated
on the day of diagnosis, thus they are classification rather than prediction
problems.
Furthermore, IDSA severity scores do not correlate well with disease-related
adverse events which are a more salient outcome to prevent.
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition (@tbl-risk).
Multiplying the NNS of the OTU-based models by the estimated NNT of 
`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
and `r pragmatic_fdx_nnb` for pragmatic severity.
Thus, in a hypothetical scenario where these assumptions about fidaxomicin hold
true, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
predicted to experience a severe outcome and be treated with fidaxomicin in
order for one patient to benefit.
As the NNS values were computed at the 95th percentile of risk (where 5% of
patients screened are predicted to experience severity), these NNB values mean
that `r min_screen` to `r max_screen` total CDI patients would need to be
screened by an OTU-based prediction model in order for one patient to benefit.
For comparison, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data extracted two days after diagnosis and from a
smaller set of manually curated variables, achieving precision values of 0.417
(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
curated model at the 95th percentile of risk
[@li_using_2019;@rao_clostridium_2015]. <!-- TODO possible to find NNS on day of diagnosis for EHR model-->
Pairing the prior EHR-based model with fidaxomicin would yield
an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
one patient to benefit, although the EHR was extracted two days after diagnosis
while OTUs in this study are from stool samples collected on the day of diagnosis.
These estimates represent a proof-of-concept demonstration of the potential
value and trade-offs of deploying severity prediction models trained on
microbial factors versus EHRs to guide clinicians' treatment decisions.

@@

Permutation feature importance revealed patterns of important bacteria that
concord with the literature.
Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and
severity has been well-documented in prior studies, thus its importance and
increase in abundance for severe cases is not surprising
[@schubert_microbiome_2014;
@antharam_intestinal_2013;
@berkell_microbiota-based_2021;
@lesniak_gut_2022].
For many of the top OTUs, there is a wide range in importance.
Notably, the OTU represented by _Pseudomonas_ had wide variance in importance
for the full dataset in models predicting attributable severity, with the
maximum point more important than any other OTU yet a minimum below zero.
However, for the intersection dataset, this OTU was removed due to having
near-zero variance.
The presence of _Pseudomonas_ was thus informative in a small number of 
patient samples, but not in others, and these samples were lost in the
intersection dataset.
Overall the abundance data are sparse, as these patients were likely on
antibiotics prior to CDI onset and may have begun antibiotic treatment prior to
the initial stool sample collection.
A limitation of permutation importance is that the contribution of each feature
is considered in isolation, but members of microbial communities interact and
compete with each other, thus these complicated relationships are not well 
captured by permutation importance.

The full pragmatic severity model performed just as well as a prior
EHR-based model trained on the day of diagnosis, demonstrating the potential
utility of OTU-based models.
In terms of the number needed to screen, the OTU-based pragmatic severity model 
outperformed a prior model of manually curated clinical variables, but not a
model trained on EHR data extracted two days after the diagnosis.
While the attributable definition had a worse NNS for the OTU-based models, it
did not perform worse than the prior curated model, and it may be the most
clinically relevant as physician chart review increases confidence that
positively-labelled severe outcomes are due to the CDI rather than other causes.
Obtaining EHR data for the dataset in this study would allow a more direct
comparison of the performance of models trained on OTUs, EHRs, or both.

However, it is not enough for models to perform well to justify deploying them
in a clinical setting; benefit over current practices must be shown [@wiens_no_2019].
Although no known treatment options are proven to reduce the risk of severe CDI
outcomes, fidaxomicin is promising due to its improved time to resolution and
reduced recurrence.
Despite its increased cost, fidaxomicin is also attractive as a preferential
antibiotic option as vancomycin-resistant enterococcus is on the rise and
enterococci are known to worsen CDI
[@poduval_clostridium_2000;@smith_enterococci_2022].
We extended our analysis of clinical value to incorporate the number needed to
treat for fidaxomicin alongside the predictive models in order to calculate the
number needed to benefit.
The NNB contextualizes model performance within clinical reality, as it combines
both model performance and treatment effectiveness [@liu_number_2019].
A more robust analysis of clinical value would further consider the cost of
treatment options versus the savings of averting severe outcomes, as economic
disparities are a major barrier to treatment in the US [@johnson_clinical_2021].
Cost-benefit analyses based on clinical trial data have reported that
fidaxomicin may be as cost-effective as vancomycin as a treatment for initial
CDI cases, largely due to the reduced risk of recurrence
[@jiang_budget_2022;@reveles_fidaxomicin_2017].
While our analysis of clinical value is only a proof-of-concept, if evidence
emerges that new or existing treatments significantly reduce the risk of severe
CDI, our results can be incorporated into future considerations of whether to
build severity prediction models and what features should be incorporated.
In practice, EHR-based models are less costly to deploy than OTU-based models
and do not require additional clinical sample collection.
Amplicon sequencing is not typically performed for CDI patients, however,
routinely profiling the microbial communities of CDI patients could be justified
if models that incorporate microbial features were shown to improve patient
outcomes.

@@

## Sample collection

@@

## 16S rRNA gene amplicon sequencing

@@

All sequences were processed with mothur (v1.46) using the MiSeq SOP protocol
[@schloss_introducing_2009; @kozich_development_2013].
Paired sequencing reads were combined and aligned with the SILVA (v132)
reference database [@quast_silva_2013] and taxonomy was assigned with a modified
version of the Ribosomal Database Project reference sequences (v16)
[@cole_ribosomal_2014].
Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
mothur [@westcott_opticlust_2017], resulting in `r num_otus` OTUs.
Only the first CDI sample per patient was used for subsequent ML analyses such
that no patient is represented more than once, resulting in a dataset of
`r n_cases_first` samples.

## Defining CDI severity

@@

## Model training

@@

## Model evaluation

@@

## Number needed to benefit

@@

## Code availability

@@

## Data availability

@@

>>>>

title: "Predicting Severity of _C. difficile_ Infections from the Taxonomic Composition of the Gut Microbiome"

@@

csl: asm.csl

@@

  We investigated whether machine learning (ML) models trained on 
  16S rRNA gene amplicon sequences from gut microbiota extracted
  from 1,277 patient stool samples on the day of CDI diagnosis could predict 
  which CDI cases led to severe outcomes.

@@

  and otherwise uses the all-cause definition.

@@

  model performance, and increased relative abundance of _Enterococcus_ was
  associated with severe outcomes.

@@

_Clostridoides difficile_ infection (CDI) is the most common nosocomial
infection in the United States, and community-acquired cases are on the rise
[@magill_multistate_2014;@feuerstadt_burden_2023].
The classic CDI case typically occurs soon after antibiotic use, which perturbs the
protective gut microbiota and allows _C. difficile_ to proliferate [@kelly_can_2012].
Non-antibiotic medications including proton-pump inhibitors and osmotic laxatives
have also been associated with increased CDI susceptibilty and inhibited clearance
[@janarthanan_clostridium_2012;@tomkovich_osmotic_2021].
Diarrhea is the primary symptom, with some patients developing colitis, 
toxic megacolon, or requiring intensive care with an in-hospital mortality rate
of approximately 8-9% [@lucado_clostridium_2012;@kassam_clostridium_2016].
Furthermore, 5-20% of initial cases reoccur within 2-8 weeks, and recurrent
cases are associated with increased morbidity and mortality risk
[@napolitano_clostridium_2017;@kelly_can_2012].
Patient risk factors for CDI-related morbidity and mortality include age greater
than 65 years, history of recurrent CDI, and co-morbid chronic illnesses [@ressler_defining_2021].
CDI remains a significant burden on the US health care system with approximately
500,000 cases annually [@guh_trends_2020;@kwon_morbidity_2015].

There is a need for robust, accurate methods to identify patients at risk of
severe CDI outcomes.
When paired with treatment options that may reduce risk of severity, prediction
models can guide clinician decision-making to improve patient outcomes while
minimizing harms from unnecessary treatment.
Numerous scoring systems for predicting severe CDI outcomes based on patient
clinical factors have been developed, but none have validated well to external
datasets nor are any in use in routine clinical practice
[@chakra_prediction_2012;@perry_external_2022].
Rather than relying on limited sets of human-curated variables,
machine learning (ML) is a promising approach that allows for use of thousands
of features to classify samples and predict outcomes.
Indeed, ML models trained on entire Electronic Health Record (EHR) data
have demonstrated improved performance over curated models
[@rao_clostridium_2015;@li_using_2019].

@@

## CDI severity.

There is not currently a consensus definition of CDI severity.
Some scoring systems leverage clinical data available during the course of CDI,
while others focus on adverse outcomes of CDI at 30 days after diagnosis
[@ressler_defining_2021;@dieterle_systemic_2020].
We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
The Infectious Diseases Society of America (IDSA) definition of severe CDI is
based on laboratory values collected on the day of diagnosis, with a case being
severe if serum creatinine level is greater than or equal to $1.5 mg/dL$ and the
white blood cell count is greater than or equal to $15 k/\mu L$
[@mcdonald_clinical_2018].
Although data for the IDSA score is straightforward to collect, it is known to
be a poor predictor of adverse outcomes [@stevens_validation_2020].
The remaining definitions we employed focus on the occurrence of adverse
outcomes, which may be more clinically relevant.
The "attributable" severity definition is based on disease-related complications
defined by the Centers for Disease Control and Prevention, where an adverse
event of ICU admission, colectomy, or death occurs within 30 days of CDI
diagnosis, and the adverse event is determined to be attributable to the CDI by
physician chart review [@mcdonald_recommendations_2007].
However, physician chart review is time-consuming and has not been completed for
all cases (n=`r attrib_labelled` out of `r allcause_yes` cases with an adverse
outcome), so we defined "all-cause" severity where a case is severe if an
adverse event occurs within 30 days of the diagnosis regardless of the cause of
the adverse event.
Finally, we defined a "pragmatic" severity definition that makes use of the
attributable definition when available and uses the all-cause definition when
chart review has not been completed, allowing us to use as many samples as we
have available while taking physicians' expert opinions into account where
possible (@fig-flowchart B).
We trained ML models to classify (in the case of the IDSA definition) or predict
(in the case of the three other definitions) severity and determined how well
OTU-based models perform for each definition.

@@

```{r sample_counts}
#| message: false
#| label: tbl-counts
#| tbl-pos: h
#| tbl-cap: "**Sample counts and proportion of severe cases.** Each severity 
#|   definition has a different number of patient samples available, as well as
#|   a different proportion of cases labelled as severe.
#| "
#| tbl-subcap: 
#|   - "Full datasets"
#|   - "Intersection of samples with all labels available"
#| layout-ncol: 2
format_table <- function(tab) {
  tab %>%
    kable(format.args = list(big.mark = ',')) %>% 
    kable_styling(font_size = table_font_size)
}
read_csv(here('results','count_table_full.csv')) %>% 
  format_table()
read_csv(here('results','count_table_int.csv')) %>% 
  format_table()
```

## Model performance.

We first set out to train the best models possible for each severity definition.
Not all samples have outcomes available for all four severity definitions due to
missing data for some patient lab values and incomplete chart review
(@fig-flowchart B), thus each severity definition had a different number of
samples when using as many samples as possible (@tbl-counts A).
We referred to these as the full datasets.
Random forest models were trained on 100 splits of the datasets into training
and test sets, and performance was evaluated on the held-out test set using the
area under the receiver-operator characteristic curve (AUROC).
Since the severity classes were highly imbalanced with different proportions of
severe samples between definitions, we also calculated the balanced precision
and the area under the balanced precision-recall curve (AUBPRC) as first
proposed by @wu_improved_2021 to describe the precision that would be expected
if the outcome classes were balanced.

@@

After training on the full datasets, the performance as measured by the AUROCs
of the training set cross-validation folds were similar to those of the held-out
test sets, indicating that the models are neither overfit nor underfit
(@fig-performance A).
As measured by AUROC on the held-out test sets, models predicting pragmatic
severity performed best with a median AUROC of `r perf_medians$auroc_full_pragmatic`,
and this was significantly different from that of the other definitions 
on the full datasets (P < 0.05).
Models predicting IDSA, all-cause, and attributable severity performed similarly
with median test set AUROCs of
`r perf_medians$auroc_full_idsa`,
`r perf_medians$auroc_full_allcause`, and
`r perf_medians$auroc_full_attrib` respectively.
The test set AUROCs were not significantly different (P > 0.05) for attributable
and IDSA nor for attributable and all-cause, but the IDSA and all-cause AUROCs
were significantly different from each other (P < 0.05).
We plotted the receiver-operator characteristic curve and found that the
pragmatic severity models outperformed the others at all specificity values
(@fig-performance B).
For comparison, a prior study with a different dataset trained a logistic
regression model on whole Electronic Health Record data extracted on the day of
CDI diagnosis to predict attributable severity, yielding an AUROC of 
`r ehr_auroc` [@li_using_2019].
While our attributable severity model did not meet this performance, the
pragmatic severity model performed just as well as the EHR-based model in terms
of AUROC.

::: {#fig-performance fig-cap="**Performance of ML models.**
In the left facets, models were trained on the full datasets, with different
numbers of samples available for each severity definition.
In the right facets, models were trained on the same dataset consisting of the
intersection of samples with labels available for all definitions.
Note that the intersection dataset has exactly the same labels for attributable
and pragmatic severity, thus these have identical performance.
**A)** 
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets, and the area under the balanced
precision-recall curve (AUBPRC) for the test sets.
Each point is annotated with the median performance across 100 train/test splits
with tails as the 95% CI.
**B)** Receiver-operator characteristic curves for the test sets.
Mean specificity is reported at each sensitivity value, 
with ribbons as the 95% CI.
**C)** Balanced precision-recall curves for the test sets.
Mean balanced precision is reported at each recall (sensitivity) value, 
with ribbons as the 95% CI.
Original unbalanced precision-recall curves are shown in Supplementary @fig-prc.
"}
![](figures/ml-performance.png)
:::

@@

## Feature importance.

@@

## Estimating clinical value.

Even if a model performs well, it may not be useful in a clinical setting unless
it can guide clinicians to choose between treatment options.
At this time, we are not aware of any direct evidence that a particular
treatment reduces the risk of severe CDI outcomes.
However, with some assumptions we offer a proof-of-concept analysis of the
potential clinical value of OTU-based severity prediction models when paired
with treatments that may reduce severity.
When considering the suitability of a model for deployment in clinical settings,
the number needed to screen (NNS) is a highly relevant metric representing how
many patients must be predicted as severe by the model to identify one true
positive.
Similarly, the number needed to treat (NNT) is the number of true positive
patients that must be treated by an intervention in order for one patient to
benefit from the treatment.
Multiplying NNS by NNT yields the number needed to benefit (NNB): the number
of patients predicted to have a severe outcome who then benefit from the
treatment [@liu_number_2019].
Thus the NNB pairs model performance with treatment effectiveness to estimate
the benefit of using predictive models in clinical practice, and are useful for
comparing models and performing cost-benefit analyses.
Lower values of NNS, NNT, and NNB are better, with the minimum value being 1,
as fewer patients must be screened and treated in order to benefit a single patient.

@@

To calculate a clinically-relevant NNS for these models, we computed the
confusion matrix at the 95th percentile of risk for each prediction model.
We excluded the IDSA severity models as the IDSA severity scores were calculated
on the day of diagnosis, thus they are classification rather than prediction
problems.
Furthermore, IDSA severity scores do not correlate well with disease-related
adverse events which are a more salient outcome to prevent.
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition (@tbl-risk).
Multiplying the NNS of the OTU-based models by the estimated NNT of 
`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
and `r pragmatic_fdx_nnb` for pragmatic severity.
Thus, in a hypothetical scenario where these assumptions about fidaxomicin hold
true, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
predicted to experience a severe outcome and be treated with fidaxomicin in
order for one patient to benefit, with the pragmatic severity models having
the best NNB when paired with fidaxomicin.
As the NNS values were computed at the 95th percentile of risk (where 5% of
patients screened are predicted to experience severity), these NNB values mean
that `r min_screen` to `r max_screen` total CDI patients would need to be
screened by an OTU-based prediction model and treated with fidaxomicin in order
for one patient to benefit.
For comparison, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data extracted two days after diagnosis and from a
smaller set of manually curated variables, achieving precision values of 0.417
(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
curated model at the 95th percentile of risk
[@li_using_2019;@rao_clostridium_2015]. 
Pairing the prior EHR-based model with fidaxomicin would yield
an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened and
treated for one patient to benefit.
Thus the pragmatic severity model did not outperform the EHR-based model,
although the EHR data were extracted two days after diagnosis while OTUs in this
study are from stool samples collected on the day of diagnosis.
These estimates represent a proof-of-concept demonstration of the potential
value and trade-offs of deploying severity prediction models trained on
microbial factors versus EHRs to guide clinicians' treatment decisions.
<!-- 
TODO possible to find NNS on day of diagnosis for EHR model?
-->

@@

Permutation feature importance revealed patterns of important bacteria that
concord with the literature.
Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and
severity has been well-documented in prior studies, thus its importance and
increase in abundance for severe cases is not surprising
[@schubert_microbiome_2014;
@antharam_intestinal_2013;
@berkell_microbiota-based_2021;
@lesniak_gut_2022].
For many of the top OTUs, there is a wide range in importance.
Notably, the OTU represented by _Pseudomonas_ had wide variance in importance
for the full dataset in models predicting attributable severity, with the
maximum point more important than any other OTU yet a minimum below zero.
However, for the intersection dataset, this OTU was removed due to having
near-zero variance.
The presence of _Pseudomonas_ was thus informative in a small number of 
patient samples, but not in others, and these samples were lost in the
intersection dataset.
Overall the abundance data are patchy, as these patients were likely all on
antibiotics for unrelated infections prior to CDI onset.
A limitation of permutation importance is that the contribution of each feature
is considered in isolation, but members of microbial communities interact and
compete with each other, thus these complicated relationships are not well 
captured by permutation importance.

The full pragmatic severity model performed just as well as a prior
EHR-based model trained on the day of diagnosis, demonstrating the potential
utility of OTU-based models.
In terms of the number needed to screen, the OTU-based pragmatic severity model 
outperformed a prior model of manually curated clinical variables, but not a
model trained on EHR data extracted two days after diagnosis.
While the attributable definition had a worse NNS for the OTU-based models, it
did not perform worse than the prior curated model, and it may be the most
clinically relevant as physician chart review increases confidence that
positively-labelled severe outcomes are due to the CDI rather than other causes.
Obtaining EHR data for the dataset in this study would allow a more direct
comparison of the performance of models trained on OTUs, EHRs, or both, as well
as extracting EHR data on the day of diagnosis rather than two days after.

However, it is not enough for models to perform well to justify deploying them
in a clinical setting; benefit over current practices must be shown [@wiens_no_2019].
Although no known treatment options have been shown to reduce the risk of severe
CDI outcomes, fidaxomicin is promising due to its improved time to resolution
and reduced recurrence.
Despite its increased cost, fidaxomicin is also attractive as a preferential
antibiotic option as vancomycin-resistant _Enterococcus_ is on the rise and
enterococci are known to worsen CDI
[@poduval_clostridium_2000;@smith_enterococci_2022].
We extended our analysis of clinical value to incorporate the number needed to
treat for fidaxomicin alongside the predictive models in order to calculate the
number needed to benefit.
The NNB contextualizes model performance within clinical reality, as it combines
both model performance and treatment effectiveness [@liu_number_2019].
A more robust analysis of clinical value would further consider the cost of
treatment options versus the savings of averting severe outcomes, as economic
disparities are a major barrier to treatment in the US [@johnson_clinical_2021].
Cost-benefit analyses based on clinical trial data have reported that
fidaxomicin may be as cost-effective as vancomycin as a treatment for initial
CDI cases, largely due to the reduced risk of recurrence
[@jiang_budget_2022;@reveles_fidaxomicin_2017].
While our analysis of clinical value is only a proof-of-concept, if evidence
emerges that new or existing treatments significantly reduce the risk of severe
CDI, our results can be incorporated into future considerations of whether to
build severity prediction models and what features should be incorporated.
In practice, EHR-based models are less costly to deploy than OTU-based models
and do not require additional clinical sample collection.
Amplicon sequencing is not typically performed for CDI patients, however,
routinely profiling the microbial communities of CDI patients could be justified
if models that incorporate microbial features were shown to improve patient
outcomes.

@@

## Sample collection.

@@

## 16S rRNA gene amplicon sequencing.

@@

All sequences were processed with mothur (v1.46) using the MiSeq SOP protocol
[@schloss_introducing_2009; @kozich_development_2013].
Paired sequencing reads were combined and aligned with the SILVA (v132)
reference database [@quast_silva_2013] and taxonomy was assigned with a modified
version of the Ribosomal Database Project reference sequences (v16)
[@cole_ribosomal_2014].
Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
mothur [@westcott_opticlust_2017], resulting in `r num_otus` OTUs.
Samples were then subsampled to 5,000 sequences per sample.
Only the first CDI sample per patient was used for subsequent ML analyses such
that no patient is represented more than once, resulting in a dataset of
`r n_cases_first` samples.

## Defining CDI severity.

@@

## Model training.

@@

## Model evaluation.

@@

## Number needed to benefit.

@@

## Code availability.

@@

## Data availability.

@@

# Supplement

<!-- TODO how get supplement figures labelled with S prefix
-->

::: {#fig-prc fig-cap="**Precision-recall curves.**
The original precision-recall curves for each model.
The horizontal line is the baseline precision, i.e. the proportion of severe
cases in the dataset for each severity definition.
Since each definition has a different baseline precision, the PRCs cannot be
compared directly without balancing the precision (see Figure 2).
"}
![](figures/prc_curves.png)
:::

::: {#fig-cdiff fig-cap="**_C. difficile_ relative abundance and feature importance.**
Of the 45 OTUs belonging to the _Peptostreptococcaceae_ family, only one
(OTU 25) had abundance values above the limit of detection.
**Left**: log~10~-transformed relative abundance of OTU 25 in the full datasets.
The dashed line is the limit of detection.
**Right**: Permutation feature importance as measured by AUROC for OTU 25.
The point is the mean difference in AUROC and the tails are the 75% confidence
interval.
The dotted line is a feature importance of zero, meaning the feature is not
important.
"}
![](figures/cdiff-otu.png)
:::

<!-- TODO supplementary figure with alpha and beta diversity & significance.
Samples were rarefied to 5,000 sequences per sample, repeated 1,000
times for alpha and beta diversity analysis.
 -->

<!--
TODO move aubprc to supplement?
-->
